---
title: "p8130_hw3_ps3194"
author: "Pangsibo Shen"
date: "10/20/2020"
output:
  html_document:
    toc: TRUE
  
---

```{r, message = FALSE}
library(tidyverse)
```
### **Problem 1**

```{r import and tidy data, message = FALSE}
exercise = read_csv("./data/Exercise.csv") %>%
  janitor::clean_names()
```

#### **a) Perform appropriate tests to assess if the Systolic BP at 6 months is significantly different from the baseline values for each of the groups:**

##### **i) Intervention group**

solution: 
we are using two-sided paired t-test $\mu_b$ stands for the mean systolic BP for baseline group and $\mu_a$ stands for the mean systolic BP for 6 month group.
The null and alternative hypotheeses are:
$$H_0: \mu_{a}- \mu_b = 0$$
$$H_1: \mu_{a}- \mu_b \neq 0$$
with a pre-specified significance level $\alpha$ = 0.05, compute the test statistics:
$$t= \frac{\bar{d} -0}{s_d/\sqrt{n}}, DF: t_{n-1} $$
where $\bar{d}$ stands for mean difference from sample and $s_d$ stands for standard deviation of differences.

```{r intervention}
intervention = exercise %>%
  filter(group == 1)

systolic_pre_1 = pull(intervention, systolic_pre) #baseline systolic data for intervention group

systolic_post_1 = pull(intervention, systolic_post) #6 month systolic data for intervention group

systolic_diff_1 = systolic_post_1 - systolic_pre_1 # difference in systolic BP

sd_diff_1 = sd(systolic_diff_1) # standard deviation for difference in systolic BP

test_systolic_1 = mean(systolic_diff_1)/(sd_diff_1/sqrt(length(systolic_diff_1))) 

#t.test(systolic_post_1, systolic_pre_1 , paired = T, alternative = "two.sided")
```

```{r}
abs(test_systolic_1) # absolute value of t statistic

qt(1 - 0.05/2,length(systolic_diff_1) - 1) # critical value
```
$$|t| > t_{n-1, 1- \alpha/2} $$
the absolute t statistic equals `r abs(test_systolic_1)` which is greater the critical value `r qt(1 - 0.05/2,length(systolic_diff_1) - 1)`. Hence we reject the null hypothesis $H_0$ which means that the Systolic BP at 6 months is significantly different from the baseline values for intervention group.

##### **ii) Control group**

solution: 
we are using two-sided paired t-test $\mu_b$ stands for the mean systolic BP for baseline group and $\mu_a$ stands for the mean systolic BP for 6 month group.
The null and alternative hypotheeses are:
$$H_0: \mu_{a}- \mu_b = 0$$
$$H_1: \mu_{a}- \mu_b \neq 0$$
with a pre-specified significance level $\alpha$ = 0.05, compute the test statistics:
$$t= \frac{\bar{d} -0}{s_d/\sqrt{n}}, DF: t_{n-1} $$
where $\bar{d}$ stands for mean difference from sample and $s_d$ stands for standard deviation of differences.

```{r control}
control = exercise %>%
  filter(group == 0)

systolic_pre_0 = pull(control, systolic_pre) #baseline systolic data for control group

systolic_post_0 = pull(control, systolic_post) #6 month systolic data for control group

systolic_diff_0 = systolic_post_0 - systolic_pre_0 # difference in systolic BP

sd_diff_0 = sd(systolic_diff_0) # standard deviation for difference in systolic BP

test_systolic_0 = mean(systolic_diff_0)/(sd_diff_0/sqrt(length(systolic_diff_0))) 

#t.test(systolic_post_0, systolic_pre_0 , paired = T, alternative = "two.sided")
```

```{r}
abs(test_systolic_0) # absolute value of t statistic

qt(1 - 0.05/2,length(systolic_diff_0) - 1) # critical value
```
$$|t| < t_{n-1, 1- \alpha/2} $$
the absolute t statistic equals `r abs(test_systolic_0)` which is less the critical value `r qt(1 - 0.05/2,length(systolic_diff_0) - 1)`. Hence we fail to reject null hypothesis $H_0$ which means that there is no difference between systolic BP at 6 month and at baseline for control group.

---------------------

#### **b)  Now perform a test and provide the 95% confidence interval to assess the Systolic BP absolute changes between the two groups.**

solution: 
We are going to use two-sample independent t-test for the problem. First, we need to test for equality of variances. 
$$H_0: \sigma_{c}^2- \sigma_i^2 = 0$$
$$H_1: \sigma_{c}^2- \sigma_i^2 \neq 0$$
with a pre-specified significance level $\alpha$ = 0.05, compute the test statistics:
$$F= \frac{s_c^2}{s_i^2}, DF: F_{n_c-1,n_i-1} $$
where $s_c^2$ stands for sample variance for control group and $s_i^2$ stands for sample variance for intervention group.
```{r variance equality test}
# Testing equality of variances for two independent samples
# drawn from two underlying normal distributions.

# control(1): s1=14.81, n1=36, x1_bar=-3.33
# intervention(2): s2=17.17, n2=36, x2_bar=-8.58

F_test = 14.81^2/17.17^2
F_test

F_crit = qf(.975, df1=35, df2=35) 
F_crit
```
$$F <  F_{n_c-1,n_i-1,1-\alpha/2} $$
F statistic equals `r F_test` which is less the critical value `r qf(.975, df1=35, df2=35)`. Hence we fail to reject the null hypothesis $H_0$ which means that there is no significant difference in pop. variance between control and intervention group.Therefore, We are going to use two-sample independent t-test with equal variances.
$$H_0: \mu_{c}= \mu_i$$
$$H_1: \mu_{c} \neq \mu_i$$
with a pre-specified significance level $\alpha$ = 0.05, compute the test statistics:
$$t= \frac{\bar{x_c} -\bar{x_i}}{s\sqrt{\frac{1}{n_c}+\frac{1}{n_i}}}, DF: t_{n_c+n_i-2} $$
where $\bar{x_c}$ stands for Systolic BP absolute changes for control group and $\bar{x_i}$ stands for Systolic BP absolute changes for intervention group.$s$ stands for pooled estimate of standard deviation which can be calculated by the formula below.

$$s = \sqrt\frac{(n_c-1)s_c^2+(n_i-1)s_i^2}{n_c+n_i-2} $$

```{r t test with equal variance}
std_pooled = sqrt(((14.81^2*35)+(17.17^2*35))/70)

t_stats = (-3.33+8.58)/(std_pooled*sqrt((1/36)+(1/36)))
t_stats
# Compare t_stats to the critical value: t with 18 df

qt(0.975,70) # 2.10

```
$$|t| < t_{n_c+n_i-2,1-\alpha/2}$$
t statistic equals `r t_stats` which is less the critical value `r qt(0.975,70)`. Hence we fail to reject the null hypothesis $H_0$ which means that there is no significant difference in the Systolic BP
absolute changes between two groups.

The formula for two sided 95% confidence interval for two-independent samples with equal variance:
$$(\bar{x_c} -\bar{x_i} - t_{n_c+n_i-2,1-\alpha/2}*s*\sqrt{\frac{1}{n_c}+\frac{1}{n_i}},\bar{x_c} -\bar{x_i} + t_{n_c+n_i-2,1-\alpha/2}*s*\sqrt{\frac{1}{n_c}+\frac{1}{n_i}})$$

```{r}
CI_lower = -3.33+8.58 - qt(0.975,70)*std_pooled*sqrt((1/36) + (1/36))
CI_lower
CI_upper = -3.33+8.58 + qt(0.975,70)*std_pooled*sqrt((1/36) + (1/36))
CI_upper
```
Hence the two sided 95% confidence interval is (`r CI_lower`,`r CI_upper`)






